---
title: "Impact of Federal Party Leadership Changes on Donation Patterns"
subtitle: "An Analysis of Political Donations in Canada 2013 - 2024"
author: 
  - Maria Mangru
thanks: "Code and data are available at: [https://github.com/MariaMangru/Ontario_Political_Donors](https://github.com/MariaMangru/Ontario_Political_Donors)."
date: today
date-format: long
abstract: "This study investigates how the ruling political parties in Canada's federal and Ontario provincial governments affect the donation behaviors of individual contributors. By analyzing donation records from 2013 to 2024, the research explores whether a party being in power influences the total amount of financial support it receives. The findings reveal that opposition parties often receive more donations than those in power, highlighting donor preferences to support challengers. This insight enhances our understanding of political financing and can help parties and policymakers develop more effective fundraising strategies."
format: pdf
number-sections: true
bibliography: references.bib
editor: 
  markdown: 
    wrap: 72
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(palmerpenguins)
```

# Introduction

Political donations play a critical role in supporting political
campaigns and shaping electoral competition. In Canada, both federal and
provincial elections rely heavily on contributions from individuals and
organizations, which influence the political landscape and often reflect
public support for different parties. Understanding what drives donation
patterns is essential for ensuring transparency and fairness in the
democratic process and for minimizing the risk of undue influence on
policy-making.

One key question is whether a party’s position in power affects the
amount of financial support it receives. While studies in other
countries, such as the U.S.A, have shown that incumbents often benefit
from more donations, particularly from interest groups looking to
maintain influence, there has been less focus on whether this holds true
in Canada. Research like Fouirnaies and Hall’s work (REFERENCE HERE) in
the U.S. shows that donors strategically target those in power, but how
these dynamics play out in Canadian elections remains under explored.
This paper aims to fill that gap by analyzing how a party's power status
influences donor behaviour in federal and Ontario provincial elections.

Using donation data from 2013 to 2024, this study compares the financial
support received by parties in power versus those in opposition. The
analysis examines total donations by year and party while accounting for
factors like election timing and party size (major party or not). By
identifying trends in how donors respond to shifts in political power,
this research sheds light on the financial strategies behind Canada’s
democratic processes.

The results show that opposition parties tend to receive higher
donations than those in power, suggesting that donors often prioritize
supporting challengers, possibly to influence political change. These
findings align with trends observed in other political systems but
highlight unique patterns in Canada. This information is valuable not
only for political parties developing fundraising strategies but also
for policymakers and researchers seeking to ensure fairness and
accountability in political financing.

The paper is organized as follows: Section \ref{sec-data} discusses the
data sources and methodology, followed by an explanation of the
variables used in the analysis. The regression models and results are
then presented, along with interpretations. The paper concludes by
discussing the broader implications of the findings and suggesting
directions for future research.\[NEED TO UPDATE THE REFERENCES\]

# Data {#sec-data}

## Data Source

# Data

## Overview

The dataset used in this analysis comprises of political donation
records spanning from 2013 to 2024, covering both the federal level and
the province of Ontario in Canada. These records were sourced from the
Investigative Journalism Foundation (IJF), which systematically collects
and maintains comprehensive data on political donations across various
Canadian jurisdictions. The IJF gathers daily updates by monitoring
elections agency websites at the federal level and within each province
and territory, ensuring the dataset remains current and robust. Although
the IJF’s repository includes historical data dating back to 1993, this
study focuses exclusively on the 2013-2024 period to capture recent
trends and changes in political financing dynamics. \[Inlude references
here for IJF, R and the packages used.\]

Political donations in Canada are subject to legal disclosure
requirements, promoting transparency and accountability within the
democratic process. Both federal and provincial election finance laws
mandate that political parties, candidates, and associated entities
report donations exceeding specific thresholds. These disclosures vary
by jurisdiction in terms of frequency and reporting formats. For
instance, Ontario requires annual returns for general and by-elections,
leadership races, and donations to parties, candidates, and constituency
associations, all available in downloadable spreadsheet formats from
2007 onward. This regulatory framework ensures that significant
contributions are systematically documented and publicly accessible,
providing a reliable foundation for this analysis.

## Measurement

The primary objective of this analysis is to investigate the impact of a
political party’s power status on donation patterns within Canadian
federal and Ontario provincial elections. The dataset includes detailed
records of donations made to various recipients, including political
parties, party leadership contestants, riding associations (also known
as electoral district or constituency associations), and individual
riding candidates. Donations are tracked across different types of
electoral events, such as general elections, by-elections, and
leadership races.

Key variables within the dataset are as follows:

-   **Political Party:** Categorical variable indicating the recipient
    political party, encompassing both major parties (e.g., Liberal
    Party, Conservative Party, New Democratic Party) and minor parties.
-   **Donation Year:** Numerical variable representing the year the
    donation was made, ranging from 2013 to 2024. This captures
    different political climates and election periods.
-   **Amount Donated:** Continuous variable denoting the monetary value
    of each donation, varying widely due to differing donor capacities
    and legal contribution limits.

### Constructed Variables

In addition to the original variables, several constructed variables
were created to facilitate a more nuanced analysis:

-   **Recipient_in_Power:** Binary variable (1/0) indicating whether the
    recipient party was in power during the donation year.
-   **Party_Size:** Binary variable (1/0) categorizing parties as
    major (1) or minor (0), based on their representation in the
    legislature.
-   **Election_Year:** Binary variable (1/0) indicating whether the
    donation was made during an election year.
-   **Log_Total_Donations:** Logarithm of the total donations received
    by a party in a given year, calculated as `log(Total_Donations + 1)`
    to handle zero values.
-   **Interaction Terms:** Interaction terms such as
    `Recipient_in_Power * Party_Size` to explore how the effect of being
    in power varies between major and minor parties.

## Data Cleaning

The raw donation records underwent a series of preprocessing steps to
ensure the dataset's integrity and suitability for analysis. The key
data cleaning procedures included:

1.  **Handling Missing Values:** Missing or incomplete records were
    addressed by either imputing values based on available information
    or excluding incomplete entries to maintain data integrity.
2.  **Standardizing Party Names:** Political party names were
    standardized across jurisdictions to ensure consistency. For
    example, variations like “Progressive Conservative,” “Progressive
    Conservative Party of Ontario,” and “The Progressive Conservative
    Association of Ontario” were consolidated into “Progressive
    Conservative Association of Ontario.”
3.  **Removing Outliers:** Extreme donation amounts that could skew the
    analysis were identified and removed based on predefined criteria.
4.  **Aggregating Donations:** Donations were aggregated to reflect
    total amounts per party per year, facilitating a higher-level
    analysis of donation patterns.
5.  **Creating Binary Variables:** Binary indicators were generated for
    key variables such as `Recipient_in_Power`, `Party_Size`, and
    `Election_Year` to streamline regression analysis.

Additionally, data from jurisdictions with incomplete historical records
or inconsistent reporting formats were excluded to maintain the
dataset's uniformity. Detailed data cleaning steps, including the
handling of OCR-converted records and manual verification processes, are
elaborated in [Appendix A: Data Cleaning Notes](#sec-appx).

```{r}
#| include: false
#| warning: false
#| message: false

# Load necessary libraries
library(tidyverse)
library(janitor)
library(knitr)
library(kableExtra)

# Read in the cleaned donation data
analysis_data <- read_csv("../data/02-analysis_data/analysis_data.csv")
```

## Outcome Variables

The main outcome variable of interest in this analysis is the **Amount
Donated**, which represents the monetary value of each donation made to
a political party within a given year. This continuous variable is
crucial for assessing the financial support received by parties based on
their power status and other influencing factors. Understanding donation
amounts helps in evaluating the extent to which financial backing
correlates with political influence and success.

```{r}
#| label: tbl-summary-ontario
#| tbl-cap: "Summary Statistics for Ontario Donations"
#| echo: false
#| message: false

#### Donations by Year ####
donations_by_year_ontario <- ontario_data %>%
  group_by(donation_year) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )

kable(donations_by_year_ontario, caption = "Summary Statistics for Ontario Donations")
```

[NEED TO INCLUDE OTHER TABLES HERE FOR ONTARIO AND ALSO FEDERAL]



## Predictor Variables

Several key predictor variables were identified to evaluate the factors
influencing donation amounts effectively:

-   **Recipient_in_Power:** Indicates whether the recipient party was in
    power during the donation year (1) or not (0). This variable is
    pivotal in assessing whether being in power attracts more financial
    support.
-   **Party_Size:** Categorizes parties as major (1) or minor (0) based
    on their representation in the legislature. Major parties typically
    have more visibility and influence, potentially attracting more
    donations.
-   **Election_Year:** Indicates whether the donation was made during an
    election year (1) or not (0). Election cycles often see increased
    financial activity as parties ramp up fundraising efforts.

### Additional Predictor Variables

-   **Log_Total_Donations:** The logarithm of total donations received
    by a party in a given year, calculated to normalize the distribution
    and handle skewness in donation amounts.
-   **Interaction Terms:** Terms such as
    `Recipient_in_Power * Party_Size` to explore the combined effect of
    power status and party size on donation amounts.

```{r}
#| include: false
#| warning: false
#| message: false

# Create log_total_donations
donations_cleaned <- donations_cleaned %>%
  group_by(party, year) %>%
  mutate(total_donations = sum(amount_donated, na.rm = TRUE)) %>%
  ungroup() %>%
  mutate(log_total_donations = log(total_donations + 1))
```

## Summary Statistics

### Ontario Data

#### Total Donations and Amount Donated

[TABLE NEEDED HERE]

#### Donations by Year

```{r}
#| include: false
#| warning: false
#| message: false

# Aggregate donations by year for Ontario
donations_by_year_ontario <- ontario_data %>%
  group_by(donation_year) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )
```

```{r}
#| label: tbl-donations-year-ontario
#| tbl-cap: "Donations by Year in Ontario"
#| echo: false
#| message: false

kable(donations_by_year_ontario, caption = "Donations by Year in Ontario")
```

#### Donations by Political Party

```{r}
#| include: false
#| warning: false
#| message: false

# Aggregate donations by party for Ontario
donations_in_power_ontario <- ontario_data %>%
  group_by(recipient_in_power) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )
```

```{r}
#| label: tbl-donations-party-ontario
#| tbl-cap: "Donations by Political Party in Ontario"
#| echo: false
#| message: false

kable(donations_in_power_ontario, caption = "Donations by Political Party in Ontario")
```

#### Donations Based on Recipient Power Status

[TABLE NEEDED HERE]

### Federal Data

#### Total Donations and Amount Donated

[TABLE NEEDED HERE]

#### Donations by Year

```{r}
#| include: false
#| warning: false
#| message: false

# Aggregate donations by year for Federal
donations_by_year_federal <- federal_data %>%
  group_by(donation_year) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )

```

```{r}
#| label: tbl-donations-year-federal
#| tbl-cap: "Donations by Year at Federal Level"
#| echo: false
#| message: false

kable(donations_by_year_federal, caption = "Donations by Year at Federal Level")
```

#### Donations by Political Party

```{r}
#| include: false
#| warning: false
#| message: false

# Aggregate donations by party for Federal
donations_by_party_federal <- federal_data %>%
  group_by(political_party) %>%
  summarize(
    total_donations = n(),
    total_amount = sum(amount, na.rm = TRUE),
    avg_amount = mean(amount, na.rm = TRUE)
  )

```

```{r}
#| label: tbl-donations-party-federal
#| tbl-cap: "Donations by Political Party at Federal Level"
#| echo: false
#| message: false

kable(donations_by_party_federal, caption = "Donations by Political Party at Federal Level")
```

#### Donations Based on Recipient Power Status

[TABLE NEEDED HERE]

## Relationships Between Variables

Preliminary analysis reveals that political parties not in power receive
higher total donations on average compared to those in power, both at
the federal and provincial levels. This consistent trend suggests that
donors may strategically allocate their contributions to opposition
parties in hopes of influencing political change and promoting
accountability. Additionally, major parties consistently attract more
donations than minor parties, underscoring the importance of party
prominence and visibility in fundraising efforts.

### Donation Amounts by Recipient Power Status

```{r}
#| label: fig-recipient-power-distribution-ontario
#| fig-cap: "Donation Amounts by Recipient Power Status in Ontario"
#| echo: false
#| message: false

p4_ontario <- ggplot(donations_in_power_ontario, aes(x = recipient_in_power, y = total_amount, fill = recipient_in_power)) +
  geom_bar(stat = "identity") +
  labs(title = "Donations Based on Party Power Status - Ontario",
       x = "Recipient Party in Power", y = "Total Amount Donated ($)") +
  scale_x_discrete(labels = c("FALSE" = "Not in Power", "TRUE" = "In Power")) +
  scale_y_continuous() +
  theme_minimal() +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 0, hjust = 0.5))
```

### Donation Amounts by Political Party

#### Ontario

```{r}
#| label: fig-donations-party-ontario
#| fig-cap: "Donations by Political Party in Ontario"
#| echo: false
#| message: false

p3_ontario <- ggplot(donations_by_party_ontario, aes(x = reorder(political_party, -total_amount), y = total_amount, fill = political_party)) +
  geom_bar(stat = "identity") +
  labs(title = "Total Donations by Political Party - Ontario",
       x = "Political Party", y = "Total Amount Donated ($)") +
  scale_y_continuous() +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        legend.position = "none")
```

#### Federal

```{r}
#| label: fig-donations-party-federal
#| fig-cap: "Donations by Political Party at Federal Level"
#| echo: false
#| message: false

donations_by_party_federal %>%
  ggplot(aes(x = reorder(party, -Total_Amount_Donated), y = Total_Amount_Donated, fill = party)) +
  geom_bar(stat = "identity") +
  theme_minimal() +
  labs(
    title = "Donations by Political Party at Federal Level",
    x = "Political Party",
    y = "Total Amount Donated ($)",
    fill = "Political Party"
  ) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```


## Relationships Between Variables

The analysis indicates that political parties not in power receive
higher total donations on average compared to those in power, both at
the federal and provincial levels. This trend suggests strategic donor
behavior aimed at supporting challengers to foster political change and
accountability. Additionally, major parties attract significantly more
donations than minor parties, underscoring the importance of party
prominence in fundraising efforts.

# Model

The goal of our modelling strategy is twofold. Firstly,...

Here we briefly describe the Bayesian analysis model used to
investigate... Background details and diagnostics are included in
[Appendix -@sec-model-details].

## Model set-up

Define $y_i$ as the number of seconds that the plane remained aloft.
Then $\beta_i$ is the wing width and $\gamma_i$ is the wing length, both
measured in millimeters.

```{=tex}
\begin{align} 
y_i|\mu_i, \sigma &\sim \mbox{Normal}(\mu_i, \sigma) \\
\mu_i &= \alpha + \beta_i + \gamma_i\\
\alpha &\sim \mbox{Normal}(0, 2.5) \\
\beta &\sim \mbox{Normal}(0, 2.5) \\
\gamma &\sim \mbox{Normal}(0, 2.5) \\
\sigma &\sim \mbox{Exponential}(1)
\end{align}
```
We run the model in R [@citeR] using the `rstanarm` package of
@rstanarm. We use the default priors from `rstanarm`.

### Model justification

We expect a positive relationship between the size of the wings and time
spent aloft. In particular...

We can use maths by including latex between dollar signs, for instance
$\theta$.

# Results

Our results are summarized in @tbl-modelresults.

```{r}
#| echo: false
#| eval: true
#| warning: false
#| message: false

library(rstanarm)

first_model <-
  readRDS(file = here::here("models/first_model.rds"))
```

```{r}
#| echo: false
#| eval: true
#| label: tbl-modelresults
#| tbl-cap: "Explanatory models of flight time based on wing width and wing length"
#| warning: false

modelsummary::modelsummary(
  list(
    "First model" = first_model
  ),
  statistic = "mad",
  fmt = 2
)
```

# Discussion

## First discussion point {#sec-first-point}

If my paper were 10 pages, then should be be at least 2.5 pages. The
discussion is a chance to show off what you know and what you learnt
from all this.

## Second discussion point

Please don't use these as sub-heading labels - change them to be what
your point actually is.

## Third discussion point

## Weaknesses and next steps

Weaknesses and next steps should also be included.

\newpage

\appendix

# Appendix {#sec-appx}

## Appendix A: Data Cleaning Notes

The data cleaning process was meticulously carried out to ensure the
dataset's reliability and consistency. The key steps included:

1.  **Importing Raw Data:**
    -   The raw donation data was imported using the `read_csv` function
        from the `tidyverse` package.
    -   Data from PDF sources were converted to CSV using OCR tools such
        as Adobe Export PDF and ABBYY FineReader, followed by manual
        verification to correct OCR-induced errors.
2.  **Handling Missing Values:**
    -   Missing values in critical variables (`amount_donated`,
        `recipient_in_power`, `party_size`, `election_year`) were
        addressed by excluding incomplete records to maintain data
        integrity.
3.  **Standardizing Party Names:**
    -   Political party names were standardized across all
        jurisdictions. For example, variations like “Progressive
        Conservative,” “Progressive Conservative Party of Ontario,” and
        “The Progressive Conservative Association of Ontario” were
        consolidated into “Progressive Conservative Association of
        Ontario.”
4.  **Removing Outliers:**
    -   Donation amounts exceeding three standard deviations from the
        mean were identified and removed to prevent skewing the
        analysis.
5.  **Aggregating Donations:**
    -   Donations were aggregated at the party-year level, resulting in
        a dataset where each row represents the total donations received
        by a party in a given year.
6.  **Creating Binary Variables:**
    -   Binary indicators were created for `recipient_in_power`,
        `party_size`, and `election_year` to facilitate regression
        analysis.
7.  **Log Transformation:**
    -   The `log_total_donations` variable was created by taking the
        logarithm of `Total_Donations + 1` to normalize the distribution
        and handle zero values.
8.  **Final Verification:**
    -   The cleaned dataset was cross-validated against original records
        to ensure accuracy and consistency.

# Appendix {.unnumbered}

# Additional data details

# Model details {#sec-model-details}

## Posterior predictive check

In @fig-ppcheckandposteriorvsprior-1 we implement a posterior predictive
check. This shows...

In @fig-ppcheckandposteriorvsprior-2 we compare the posterior with the
prior. This shows...

```{r}
#| eval: true
#| echo: false
#| message: false
#| warning: false
#| label: fig-ppcheckandposteriorvsprior
#| layout-ncol: 2
#| fig-cap: "Examining how the model fits, and is affected by, the data"
#| fig-subcap: ["Posterior prediction check", "Comparing the posterior with the prior"]

pp_check(first_model) +
  theme_classic() +
  theme(legend.position = "bottom")

posterior_vs_prior(first_model) +
  theme_minimal() +
  scale_color_brewer(palette = "Set1") +
  theme(legend.position = "bottom") +
  coord_flip()
```

## Diagnostics

@fig-stanareyouokay-1 is a trace plot. It shows... This suggests...

@fig-stanareyouokay-2 is a Rhat plot. It shows... This suggests...

```{r}
#| echo: false
#| eval: true
#| message: false
#| warning: false
#| label: fig-stanareyouokay
#| fig-cap: "Checking the convergence of the MCMC algorithm"
#| fig-subcap: ["Trace plot", "Rhat plot"]
#| layout-ncol: 2

plot(first_model, "trace")

plot(first_model, "rhat")
```

\newpage

# References
